---
title: "hw1_Evans"
author: "Tessa Devereaux Evans"
date: "21 February 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (ggplot2)
library (dplyr)
library (tidyr)
```


question 1
```{r}
democracy <- read.csv("democracy.csv", stringsAsFactors = FALSE, na.strings=".")
```

question 2
```{r}
#Section 2 - Initial data exploration
hist(democracy$POLLIB)
hist(democracy$GDPW, 
     main="Histogram for GDP per capita",
     xlab="GDP per capita",
     border="black",
     col="light yellow")
     
```

```{r}

hist (log10(democracy$GDPW),
     main="Histogram for log GDP per capita",
     xlab="log GDP per capita",
     border="black",
     col = "light pink")
     
```

This histogram uses a log transformation to make a skewed distribution less skewed. This can be valuable for making patterns in the data more interpretable and helping to meet the assumptions of inferential statistics. 


Question 3




Create a scatterplot of political liberties against GDP per capita.

When there is a lot of overlap in a scatter plot it is useful to “jitter” the points 

Create a scatterplot of political liberties against log GDP per capita. Jitter the points. How is the relationship different than when GDP per capita was not logged.

Create a boxplot of GDP per capita for oil producing and non-oil producing nations.

```{r}
ggplot(democracy, aes(x=democracy$GDPW, y=democracy$POLLIB)) +
  geom_point()+
  labs(title = "Civil Liberties and GDP")

ggplot(democracy, aes(x = GDPW ,y = POLLIB)) +
geom_jitter(alpha =.1) +
labs(title = "Political Liberties and GDP", y="Political Liberties", x ="GDP per Cap")

```

```{r}
ggplot(democracy, aes(x=log10(democracy$GDPW), y=democracy$POLLIB)) +
  geom_jitter(alpha=.1)+
  theme_bw()
  labs(title = "Jitter Scatterplot of Political Liberties and Log GDP per Capita")
```
  
  The Logged Scatterplot is more spread out so you can see the relationship more clearly
  
  
  
```{r}  
  
boxplot(democracy$GDPW~democracy$OIL,data=democracy, main="Boxplot of Oil and GDPW", xlab="Oil Producing Countries", ylab="GDPW")
  geom_jitter(alpha=.1)+
  theme_bw()

```

###Section 4 - Transform data and analyse

1 - Calculate the mean GDP per capita in countries with at least 40 percent Catholics. How does it compare to mean GDP per capita for all countries?

The Catholic mean is higher 

```{r}
mean(democracy$GDPW)
catholic_gdpw <-filter(democracy, CATH >= .4)$GDPW %>% mean(na.rm = TRUE)
catholic_gdpw
```

2 - Calculate the average GDP per capita in countries with greater than 60% ethnolinguistic fractionalization, less than 60%, and missing ethnolinguistic fractionalization.

```{r}
ethno_gdpw <-filter(democracy, ELF60 > .6)$GDPW %>% mean(na.rm = TRUE)
ethno_gdpw

ethno_gdpw <-filter(democracy, ELF60 < .6)$GDPW %>% mean(na.rm = TRUE)
ethno_gdpw

ethno_gdpw <-filter(democracy, is.na(ELF60))$GDPW %>% mean(na.rm = TRUE)
ethno_gdpw

```

3. What was the median of the average years of education in 1985 for all countries?

Answer is 5.625 years

```{r}

democracy_median_1985 <- filter(democracy, YEAR==1985)
summary(democracy_median_1985$EDT)


```



4. Which country was (or countries were) closest to the median years of education in 1985 among all countries?

Answer is Venezuela 

```{r}

which.min(abs(democracy_median_1985$EDT-5.625))

filter (democracy_median_1985, row_number()==77L)



```



5. What was the median of the average years of education in 1985 for democracies?

Answer is 7.805 

```{r}

democracy_1985_democracy <- filter(democracy, YEAR==1985, REG==1)
summary(democracy_1985_democracy$EDT)

```


6. Which democracy was (or democracies were) closest to the median years of education in 1985 among all democracies?

Answer is Maritius 


```{r}

which.min(abs(democracy_1985_democracy$EDT-7.805))
filter (democracy_1985_democracy, row_number()==1)

```


7. What were the 25th and 75th percentiles of ethnolinguistic fractionalization for new and old countries?


```{r}
quantile(democracy$ELF60, na.rm = TRUE)
```













